Arielle Childs, Gracie OBrien, and Katie Gately 
---
title: "Simulations In-Class Project"
date: "Due October 13, 2017 at 11:59pm"
output:
  html_document

---

<style type="text/css">
.table {

    width: 80%;
    margin-left:10%; 
    margin-right:10%;
}
</style>
```{r,setup, echo=FALSE, cache=TRUE}
## numbers >= 10^5 will be denoted in scientific notation,
## and rounded to 2 digits
options(scipen = 3, digits = 3)
```

#Project Goals:

With this project we will simulate a famous probability problem. This will not require knowledge of probability or statistics but only the logic to follow the steps in order to simulate this problem. This is one way to solve problems by using the computer. 


Since you all have completed problem 1, you first step will be to work through each of your groupmates code for problem #1 and comment on what is happening. Then as a team move forward in on direction as you move on to the rest of the problems 2-5

 1. **Gambler's Ruin**: Suppose you have a bankroll of $1000 and make bets of $100 on a fair game. By simulating the outcome directly for at most 5000 iterations of the game (or hands), estimate:
 
    a. the probability that you have "busted" (lost all your money) by the time you have placed your one hundredth bet. 
    
```{r}
set.seed(1)
gambler1=function(bet,bankroll,prob) {
  b=0   #Hand=0
  while(bet>0) {
    ProbVal=runif(1) #generates a single normal random number
    if(ProbVal <= prob) {
      bankroll = bet + bankroll 
      bet = (bet + bankroll)*0.1
    } else {
      bankroll = bankroll-bet
      bet = bet
      }
    b=b+1
    if(bankroll < bet) { #if no money to make a bet, then break
      break
      }
  }
  duration = b
  return(duration)
}
durations = numeric()
s = 5000
for(i in 1:s) {
  dice = gambler1(100,900,0.5)
  durations = c(durations,dice) #Creating a vector of durations and dice 
}

dur1=length(durations)
bust1=length(which(durations<=100))
bust1/dur1
```
The probability that you have "busted" by the time you placed your 100th bet is 0.12.

  b. the probability that you have busted by the time you have placed your five hundredth bet by simulating the outcome directly. 
    
```{r}
dur2=length(durations)
bust2=length(which(durations<=500))
bust2/dur2
```
The probability that you have "busted" by the time you placed your 500th bet is 0.477.

  c. the mean time you go bust, given that you go bust within the first 5000 hands.

```{r}
mean_bust=durations[which(durations<=5000)]
mean(mean_bust)
```
The mean time you go bust within the first 5000 hands is 758. 

  d. the mean and variance of your bankroll after 100 hands (including busts).
    
```{r}
set.seed(1)
gambler2=function(bet,bankroll,prob) {
  b=0
  while(bet>0) {
    ProbVal=runif(1)
    if(ProbVal <= prob) {
      bankroll = bet + bankroll
      bet = (bet + bankroll)*0.1
    } else {
      bankroll = bankroll-bet
      bet = bet
      }
    b=b+1
    if(b ==100) {
      break 
      }
    if(bankroll<bet){ 
      break
      }
  }
  duration = b
  return(bankroll)
}

bankroll=numeric()
s = 5000
for(i in 1:s) {
  dur3 = gambler2(100,900,0.5)
  bankroll=c(bankroll,dur3)
}

mean(bankroll)
var(bankroll)
```
The mean of the bankroll after 100 hands is 879 and the variance is 2716298. 

  e. the mean and variance of your bankroll after 500 hands (including busts).
    
```{r}
set.seed(1)
gambler3=function(bet,bankroll,prob) {
  b=0
  while(bet>0) {
    ProbVal=runif(1)
    if(ProbVal <= prob) {
      bankroll = bet + bankroll
      bet = (bet + bankroll)*0.1
    } else {
      bankroll = bankroll-bet
      bet = bet
      }
    b=b+1
    if(b ==500) {
      break
      }
    if(bankroll< bet) {
      break
      }
  }
  duration = b
  return(bankroll)
}

bankroll=numeric()
s = 5000
for(i in 1:s)
{
  dur4 = gambler3(100,900,0.5)
  bankroll=c(bankroll,dur4)
}

mean(bankroll)
var(bankroll)
```
The mean of the bankroll after 500 hands is 536 and the variance is 19437196. 
 
Note: you *must* stop playing if your player has gone bust. How will you handle this in the `for` loop?

2. Repeat the previous problem with betting on black in American roulette, where the probability of winning on any spin is 18/38 for an even payout.

a. 
```{r}
set.seed(1)
roulette=function(bet,bankroll,prob) {
  b=0   #Hand=0
  while(bet>0) {
    ProbVal=runif(1) #generates a single normal random number
    if(ProbVal <= prob) {
      bankroll = bet + bankroll
      bet = (bet + bankroll)*0.1
    } else {
      bankroll = bankroll-bet
      bet = bet
      }
    b=b+1
    if(bankroll < bet) {
      break
      }
  }
  duration = b
  return(duration)
}
dura = numeric()
s = 5000
for(i in 1:s) {
  dice = roulette(100,900,0.474) #changed the probability to 18/38 = 0.474
  dura = c(dura,dice) #Creating a vector of durations and dice 
}

dur1=length(dura)
bust1=length(which(dura<=100))
bust1/dur1
```
The probability that you have "busted" by the time you placed your 100th bet is 0.174.

b. 
```{r}
dur2=length(dura)
bust2=length(which(dura<=500))
bust2/dur2
```
The probability that you have "busted" by the time you placed your 500th bet is 0.628.

c.
```{r}
mean_bust=dura[which(dura<=5000)]
mean(mean_bust)
```
The mean time you go bust within the first 5000 hands is 507. 

d.
```{r}
set.seed(1)
roulette2=function(bet,bankroll,prob) {
  b=0
  while(bet>0) {
    ProbVal=runif(1)
    if(ProbVal <= prob) {
      bankroll = bet + bankroll
      bet = (bet + bankroll)*0.1
    } else {
      bankroll = bankroll-bet
      bet = bet
      }
    b=b+1
    if(b ==100) {
      break 
      }
    if(bankroll<bet){ 
      break
      }
  }
  duration = b
  return(bankroll)
}

bankroll=numeric()
s = 5000
for(i in 1:s) {
  dur3 = roulette(100,900,0.474)
  bankroll=c(bankroll,dur3)
}

mean(bankroll)
var(bankroll)
```
The mean of the bankroll after 100 hands is 507 and the variance is 256115. 

e.
```{r}
set.seed(1)
roulette3=function(bet,bankroll,prob) {
  b=0
  while(bet>0) {
    ProbVal=runif(1)
    if(ProbVal <= prob) {
      bankroll = bet + bankroll
      bet = (bet + bankroll)*0.1
    } else {
      bankroll = bankroll-bet
      bet = bet
      }
    b=b+1
    if(b ==500) {
      break
      }
    if(bankroll< bet) {
      break
      }
  }
  duration = b
  return(bankroll)
}

bankroll=numeric()
s = 5000
for(i in 1:s)
{
  dur4 = roulette3(100,900,0.474)
  bankroll=c(bankroll,dur4)
}

mean(bankroll)
var(bankroll)
```
The mean of the bankroll after 100 hands is 32.2 and the variance is 59837. 

3. **Markov Chains**. Suppose you have a game where the probability of winning on your first hand is 48%; each time you win, that probability goes up by one percentage point for the next game (to a maximum of 100%, where it must stay), and each time you lose, it goes back down to 48%. Assume you cannot go bust and that the size of your wager is a constant $100.
    
    a. Is this a fair game? Simulate one hundred thousand sequential hands to determine the size of your return. Then repeat this simulation 99 more times to get a range of values to calculate the expectation.
    
```{r}
set.seed(1)                             #set seed so simulation can be reproduced
markchain=function(bets,bankroll,prob)  #create a function with 3 arguments
{
  h=0                                   #no hands
  probability<-prob                     #the probability equals the given argument
  while(bets>0) {                        #while there is an actual bet.. 
    pval=runif(1)                       #generates a normal random number
    if(pval <= prob)                    #if generated number is less than or equal to the probability...
    {
      bankroll = bankroll +bets         #define the secondary argument as the combination of the first and second given arguments
      prob = prob+0.01                  #define the tertiary argument as the given argument + 0.01 (prob increases by 1 percentage point for the next game)
    }else{                              #if pval is not <= prob, then...
      bankroll = bankroll-bets          #define the secondary argument as the amount of money minus the bet (lost the hand)
      prob = probability                #the probability will stay the same
    }
    h=h+1                               #the # of hands increases
    if(h==100000){break}                #once the # of hands reaches 100000, break
    if(bankroll < bets){break}          #once the bankroll is less than the amount of the bet, break (can't afford more bets)
  }
  return(bankroll)                      #return the new value for bankroll
}

size = numeric()                        #size is a vector
it3 = 100                               #create value for simulation iterations
for(i in 1:it3)                         #for i in the length of the iterations
{
  next3 = markchain(100,900,0.48)       #use previosly defined function to simulate 99 more times        
  size = c(size,next3)                  #define size as a vector
}
mean(size)                              #calculate the mean

```
This is not fair because the mean size is 0. 

  b. Repeat this process but change the starting probability to a new value within 2% either way. Get the expected return after 100 repetitions. Keep exploring until you have a return value that is as fair as you can make it. Can you do this automatically?
    
```{r}
set.seed(1)

increment<-function(N,v){
 output=numeric()
  for (j in 1:25){
  size=numeric()
  N=N
  for(i in 1:N)
{
  next4 = markchain(100,900,0.48+v*j)
  size = c(size,next4)
}
  next_val=length(which(size==0))/N
  output=c(output,next_val)
}
  return(output)
}

increment(N=100,0.002)
#It looks like 51 is the fairest the game will be
```
The probability increment is most fair at a value of 0.05. 

   c. Repeat again, keeping the initial probability at 48%, but this time change the probability increment to a value different from 1%. Get the expected return after 100 repetitions. Keep changing this value until you have a return value that is as fair as you can make it. 

```{r}
set.seed(1)
markchain1=function(bets,bankroll,prob,incr)
{
  h=0
  probability<-prob
  while(bets>0) {
    pval=runif(1)
    if(pval <= prob)
    {
      bankroll = bankroll +bets
      prob = prob+incr
    }else{
      bankroll = bankroll-bets
      prob = probability
    }
    h=h+1
    if(h==100000){break}
    if(bankroll < bets){break}
  }
  return(bankroll)
}

increment1<-function(N,v,g){
 output=numeric()
  for (j in 1:20){
  size=numeric()
  N=N
  for(i in 1:N)
{
  next4 = markchain1(100,900,0.48+v*j, g)
  size = c(size,next4)
}
  next_val=length(which(size==0))/N
  output=c(output,next_val)
}
  return(output)
}


increment1(N=100, 0.002, 0.025)
increment1(N=100, 0.002, 0.018)
increment1(N=100, 0.002, 0.015)
increment1(N=100, 0.002, 0.012)

#If the probability increment is changed to 3%, it looks like 51 is the fairest the game will be.
```
The probability increment is most fair at a value of 0.03.

4. Creating a Bootstrap function. There is a particular concept called [bootstrapping]
(https://en.wikipedia.org/wiki/Bootstrapping_(statistics)) where we can easily create 95% confidence intervals, even for complex estimators.

The steps of this process are:

  a. Draw a sample, with replacement, from your data which is the same length of your data. 
  b. Calculate the statistic of interest on this boostrap sample (ie mean, variance, regression,...)
  c. Peform steps 1:2 at least 1000 times over until you have a vector of your statistics. 
  d. The lower bound of a 95% CI will be the 0.025 percentile
  e. The upper bound of a 95% CI will be the 0.975 percentile
Make a function called `boot_ci` which calculates the 95% confidence interval in this manner.

```{r}
x<-1:50

#boot1<-function(x, N){    #N equals # of times to sample
  #for (i in 1:N){
  #data<-sample(x, length(x),replace = T)
  #a<-mean(data, na.rm=TRUE)
  #s<-sd(data, na.rm=TRUE)
  #n <- length(data)
  #v<-var(data, na.rm=TRUE)
  #error <- qnorm(0.975)*s/sqrt(n)
  #lowbound <- a-error
  #upbound <- a+error
#}
  #return(c(a, v, lowbound, upbound))
#}

#boot2<-function(x, N, FUN, z){    #N equals # of times to sample
  #for (i in 1:N){
  #data<-sample(x, length(x),replace = T)
  #fxn<-FUN(data)
  #a<-mean(data, na.rm=TRUE)
  #s<-sd(data, na.rm=TRUE)
  #n <- length(data)
  #v<-var(data, na.rm=TRUE)
  #error <- qnorm(z)*s/sqrt(n)
  #lowbound <- a-error
  #upbound <- a+error
#}
  #return(c(fxn, lowbound, upbound))
#}

boot_ci<-function(x, N, FUN, z){    #x=data, N=# of times to sample, FUN=function of interest, z=z-score of interest
  sample<-c()                       #initiate vector
  for (i in 1:N){                   #for each iteration
  data<-sample(x, length(x),replace = T)  #sample whole length with replacement
  sample[i]<-FUN(data)              #store function of interest from bootstrapped sample in the vector
  a<-mean(sample, na.rm=TRUE)       #Calculate mean, sd, and length to use for calculation of CI
  s<-sd(sample, na.rm=TRUE)
  n <- length(sample)
  error <- qnorm(z)*s/sqrt(n)       #use mean, sd, and length to calculate error
  lowbound <- a-error               #calculate lower bound and upper bound by subtracting or adding error
  upbound <- a+error
}
  return(c(lowbound, upbound))      #retrun the confidence intervals
}
boot_ci(x, 1000, FUN=var, 0.975)
boot_ci(x, 1000, FUN=mean, 0.975)

```

5. For problems 3b and 3c, you calculated a mean value. Because you saved these final results in a vector, use the bootstrap to estimate the variance of the return in each case for your final answer. Once you have these results, which game has the smaller variance in returns?

