---
title: "Week07-inclass-GOBrien"
output: html_document
---

---
title: "Grace O'Brien Simulations Pre-Class Project"
date: "Due March 13, 2017 at 5:00pm"
output:
  html_document


---

<style type="text/css">
.table {

    width: 80%;
    margin-left:10%; 
    margin-right:10%;
}
</style>
```{r,setup, echo=FALSE, cache=TRUE}
## numbers >= 10^5 will be denoted in scientific notation,
## and rounded to 2 digits
options(scipen = 3, digits = 3)
```




#Project Goals:


With this project we will simulate a famoues probability problem. This will not require knowledge of probability or statistics but only the logic to follow the steps in order to simulate this problem. This is one way to solve problems by using the computer. 

 1.  Suppose you have a bankroll of $1000 and make bets of $100 on a fair game. By simulating the outcome directly for at most 5000 iterations of the game (or hands), estimate:

```{r}
set.seed(1)
gamble <- function(bet,money,p){
    k=0
  while(bet>0) {
    ProbVal=runif(1)
    if(ProbVal <= p)
    {
      money = bet + money
      bet = (bet + money)*0.1
    }else{
      money = money-bet # if lost bet subtract
      bet = bet
      }
    k=k+1
    if(money < bet){break}
  }
  game = k
  return(game)
}

game = numeric() #game is number of hands/ number of bets placed. A vector. 
s = 5000
for(i in 1:s)
{
  nextd = gamble(100,900,0.5)
  game = c(game,nextd)
}

game  <- rep(NA, N)
game[1] <- 1000
N <- 5000
for(i in 1:N){
hand = game[i-1] + 200*rbinom(1,1,0.500) -100
game[i] = hand
if (game[i] == 0){
  return(game[1:i])
  break
}}
return(game)

length(game) to get number of hands
game[length(game)] tells us the last value. Use to check that the last hand is zero.
```
    a. the probability that you have "busted" (lost all your money) by the time you have placed your one hundredth bet. 
    ```{r}
a=length(game) # gives number of hands until bust
b=length(which(game<=100)) # gives number of hands until bust where number of hands until bust is less than or equal to 100. 
prob=b/a # divide b/a to get probability busted by 100th bet. Simple probability rules. 
prob
    ```
    Probability that lost all money by 100th bet is 0.12.
    b. the probability that you have busted by the time you have placed your five hundredth bet by simulating the outcome directly. 
    ```{r}
a=length(game) # similar reasoning from part A. 
b=length(which(game<=500))
prob=b/a
prob
    ```
    Probability that lost all money by 500th bet is 0.477.
    c. the mean time you go bust, given that you go bust within the first 5000 hands.
    ```{r}
a=game[which(game<=5000)] # similar reasoning from part A. 
mean=mean(a)
mean
    ```
    Mean time is 758 to bust given that you go bust within first 5000 hands. 
    d. the mean and variance of your bankroll after 100 hands (including busts).
    ```{r}
set.seed(1)
GR1=function(bet,money,p)
{
  k=0
  while(bet>0) {
    ProbVal=runif(1)
    if(ProbVal <= p)
    {
      money = bet + money
      bet = (bet + money)*0.1
    }else{
      money = money-bet
      bet = bet
      }
    k=k+1 # go to next Hand
    if(k ==100){break}
    if(money<bet){break}
  }
  game = k
  return(money)
  
}

money=numeric()
s = 5000
for(i in 1:s)
{
  nextb = GR1(100,900,0.5)
  money=c(money,nextb)
}

mean(money)
var(money)
    ```
    Mean time is 879 with variance of 2716298.
    e. the mean and variance of your bankroll after 500 hands (including busts).
    ```{r}
    set.seed(1)
GR2=function(bet,money,p)
{
  k=0
  while(bet>0) {
    ProbVal=runif(1)
    if(ProbVal <= p)
    {
      money = bet + money
      bet = (bet + money)*0.1
    }else{
      money = money-bet
      bet = bet
      }
    k=k+1
    if(k ==500){break}
    if(money< bet){break}
  }
  game = k
  return(money)
  
}

money=numeric()
s = 5000
for(i in 1:s)
{
  nextb = GR2(100,900,0.5)
  money=c(money,nextb)
}

mean(money)
var(money)
    ```
    Mean time of 536 with variance of 19437196.
Note: you *must* stop playing if your player has gone bust. How will you handle this in the `for` loop?
Add {break} statements.

2. Repeat the previous problem with betting on black in American roulette, where the probability of winning on any spin is 18/38 for an even payout.

```{r}
# I copied the code I made for question 1 and just changed the probability to 18/38 instead of 0.5000000. :)
set.seed(1)
gamble2 <- function(bet,money,p){
    k=0
  while(bet>0) {
    ProbVal=runif(1)
    if(ProbVal <= p)
    {
      money = bet + money
      bet = (bet + money)*0.1
    }else{
      money = money-bet # if lost bet subtract
      bet = bet
      }
    k=k+1
    if(money < bet){break}
  }
  game = k
  return(game)
}

game = numeric() #game is number of hands/ number of bets placed. A vector. 
s = 5000
for(i in 1:s)
{
  nextd = gamble2(100,900,18/38)
  game = c(game,nextd)
}
```
    a. the probability that you have "busted" (lost all your money) by the time you have placed your one hundredth bet. 
    ```{r}
a=length(game) # gives number of hands until bust
b=length(which(game<=100)) # gives number of hands until bust where number of hands until bust is less than or equal to 100. 
prob=b/a # divide b/a to get probability busted by 100th bet. Simple probability rules. 
prob
    ```
    Probability that lost all money by 100th bet is 0.1744.
    b. the probability that you have busted by the time you have placed your five hundredth bet by simulating the outcome directly. 
    ```{r}
a=length(game) # similar reasoning from part A. 
b=length(which(game<=500))
prob=b/a
prob
    ```
    Probability that lost all money by 500th bet is 0.6298.
    c. the mean time you go bust, given that you go bust within the first 5000 hands.
    ```{r}
a=game[which(game<=5000)] # similar reasoning from part A. 
mean=mean(a)
mean
    ```
    Mean time is 504 to bust given that you go bust within first 5000 hands. 
    d. the mean and variance of your bankroll after 100 hands (including busts).
    ```{r}
set.seed(1)
GR1=function(bet,money,p)
{
  k=0
  while(bet>0) {
    ProbVal=runif(1)
    if(ProbVal <= p)
    {
      money = bet + money
      bet = (bet + money)*0.1
    }else{
      money = money-bet
      bet = bet
      }
    k=k+1 # go to next Hand
    if(k ==100){break}
    if(money<bet){break}
  }
  game = k
  return(money)
  
}

money=numeric()
s = 5000
for(i in 1:s)
{
  nextb = GR1(100,900,18/38)
  money=c(money,nextb)
}

mean(money)
var(money)
    ```
    Mean time is 464 with variance of 864712.
    e. the mean and variance of your bankroll after 500 hands (including busts).
    ```{r}
    set.seed(1)
GR2=function(bet,money,p)
{
  k=0
  while(bet>0) {
    ProbVal=runif(1)
    if(ProbVal <= p)
    {
      money = bet + money
      bet = (bet + money)*0.1
    }else{
      money = money-bet
      bet = bet
      }
    k=k+1
    if(k ==500){break}
    if(money< bet){break}
  }
  game = k
  return(money)
  
}

money=numeric()
s = 5000
for(i in 1:s)
{
  nextb = GR2(100,900,18/38)
  money=c(money,nextb)
}

mean(money)
var(money)
    ```
    Mean time of 31 with variance of 58251.


3. **Markov Chains**. Suppose you have a game where the probability of winning on your first hand is 48%; each time you win, that probability goes up by one percentage point for the next game (to a maximum of 100%, where it must stay), and each time you lose, it goes back down to 48%. Assume you cannot go bust and that the size of your wager is a constant $100.
    a. Is this a fair game? Simulate one hundred thousand sequential hands to determine the size of your return. Then repeat this simulation 99 more times to get a range of values to calculate the expectation.
    ```{r}
    MC=function(bet,bankroll,p)
{
  k=0
  prob<-p
  while(bet>0) {
    ProbVal=runif(1)
    if(ProbVal <= p)
    {
      bankroll = bankroll +bet
      p = p+0.01
    }else{
      bankroll = bankroll-bet
      p = prob
    }
    k=k+1
    if(k==100000){break}
    if(bankroll < bet){break}
  }
  return(bankroll)
}

size = numeric()
s = 100
for(i in 1:s)
{
  nextb = MC(100,900,0.48)
  size = c(size,nextb)
}
mean(size)
    ```
    Mean size is zero. This is not a fair game. 
    b. Repeat this process but change the starting probability to a new value within 2% either way. Get the expected return after 100 repetitions. Keep exploring until you have a return value that is as fair as you can make it. Can you do this automatically?
    ```{r}
    # Trying with 0.46
set.seed(1)
proportion=numeric()
for (j in 1:20){
  size=numeric()
  s = 100
for(i in 1:s)
{
  nextb = MC(100,900,0.46+0.002*j)
  size = c(size,nextb)
}
  nextproportion=length(which(size==0))/100
  proportion=c(proportion,nextproportion)
}
proportion
    # Trying with 0.485
set.seed(1)
proportion=numeric()
for (j in 1:20){
  size=numeric()
  s = 100
for(i in 1:s)
{
  nextb = MC(100,900,0.485+0.002*j)
  size = c(size,nextb)
}
  nextproportion=length(which(size==0))/100
  proportion=c(proportion,nextproportion)
}
proportion
    ```
    0.05 starting probability returns the most fair game.
    c. Repeat again, keeping the initial probability at 48%, but this time change the probability increment to a value different from 1%. Get the expected return after 100 repetitions. Keep changing this value until you have a return value that is as fair as you can make it. 
    ```{r}
    set.seed(1)
MC1=function(bet,bankroll,p,increment)
{
  k=0
  prob<-p
  while(bet>0) {
    ProbVal=runif(1)
    if(ProbVal <= p)
    {
      bankroll = bankroll +bet
      p = p+increment
    }else{
      bankroll = bankroll-bet
      p = prob
    }
    k=k+1
    if(k==100000){break}
    if(bankroll < bet){break}
  }
  return(bankroll)
}

proportion = numeric()
for (z in 1:20){
size=numeric()
s = 100
for(i in 1:s)
{
  nextb = MC1(100,900,0.48,0.002*z)
  size = c(size,nextb)
}
nextproportion=length(which(size==0))/100
proportion=c(proportion,nextproportion)
}
proportion
    ```
    0.035 probability increment allows for the proportion to return as 0.5, the fairest possible game. 
    

4. Creating a Bootstrap function. There is a particular concept called [bootstrapping]
(https://en.wikipedia.org/wiki/Bootstrapping_(statistics)) where we can easily create 95% confidence intervals, even for complex estimators.

The steps of this process are:

  a. Draw a sample, with replacement, from your data which is the same length of your data.
```{r}
samp <- sample(10, replace = TRUE)
```
  b. Calculate the statistic of interest on this boostrap sample (ie mean, variance, regression,...)
```{r}
mean(samp)
var(samp)
```
  c. Peform steps 1:2 at least 1000 times over until you have a vector of your statistics. 
```{r}
meann <- function(data, num) {
    resamples <- lapply(1:num, function(i) sample(data, replace=T)) # why function(i)?? 
    r.mean <- sapply(resamples, mean)
    r.var <- sapply(resamples, var)
    std.err <- sqrt(var(r.mean))
    list(var = r.var, std.err=std.err, resamples=resamples, mean=r.mean)   
}
meann(samp, 10)
```
  d. The lower bound of a 95% CI will be the 0.025 percentile
    e. The upper bound of a 95% CI will be the 0.975 percentile
Make a function called `boot_ci` which calculates the 95% confidence interval in this manner. 
```{r}
boot_ci <- function(data, num) {
  data<-sample(10,replace = TRUE)
  for (i in 1:length(data)) {
  resamples <- lapply(1:num, function(i) sample(data, replace=T)) 
  a<-sapply(resamples,mean)
  s<-sapply(resamples,sd)
  return(data.frame(mean = a)
  }
}

boot_ci2 <- function(data, num){
  n <- length(data)
  error <- qnorm(0.975)*s/sqrt(n)
  left <- a-error
  right <- a+error
}
head(boot_ci2(data,1000))

```


5. For problems 3b and 3c, you calculated a mean value. Because you saved these final results in a vector, use the bootstrap to estimate the variance of the return in each case for your final answer. Once you have these results, which game has the smaller variance in returns?
```{r}
boot_ci()


```
    
